A Survey of Physical Design Techniques of Information Systems

### 2010-07-13 trunk/cs848-db/Paper.tex

% make the title area
\maketitle

@@

Database users write queries and updates using a "user-friendly" language such as SQL without having to worry about how the underlying data is stored. Physical
database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed. Data is stored in the
form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
queries are answered and executed efficiently. This is reflected in designing proper indexing techniques, partitioning data for more efficient access, designing
materialized views, data clustering, data compression, striping, mirroring and denormalization~\cite{lightstone2007physical}.

This includes which indices to use, and in which order to access the tables~\cite{finkelstein1988physical} and so on.

This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
are disk based relational databases, main memory relational databases, data warehouses, and XML databases. The aim of this survey is not to explain the details
of every data structure used in physical design, but rather to compare how the different data structures have been used in these systems. This includes
explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference on the details of the
data structures used in physical database design, please see (CITE HERE).

The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design. For example, this
includes explaining what indexes are and briefly mentioning what the different types of indexes used in information systems are. After explaining the different
elements of physical database design, Section~\ref{SEC-DIFFSYS} discusses how each of the four systems uses these elements to improve its efficiency.
Sections~\ref{SEC-DDRDB},~\ref{SEC-MMDB},~\ref{SEC-WAREHOUSES},~\ref{SEC-XML} discuss Disk Based Databases, Main Memory Databases, Data Warehouses, and XML
Databases respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
Section~\ref{SEC-AUTO} then presents physical database design can be automated. Section~\ref{SEC-OPEN} mentions some of the open problems of physical database
design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this summary.

@@

@@

B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
instead of having one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE HERE).
B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.

@@

@@

Despite the fact that materialized views provide a significant improvement in performance, we cannot simply create a materialized view for each common query. To
begin with, materialized views consume disk storage which might be limited. The second, and main, problem with materialized views is their maintenance. Ensuring
that the views have the most up to date data is tricky, and may outset the benefit of using materialized views if it is not properly designed. Finally, having
several materialized views may increase the cost of searching for the appropriate view to use during the query optimization stage. Data warehouses heavily rely
on materialized views since they have many queries with several joins. This will be discussed in details in Section~\ref{SEC-WAREHOUSES}.

@@

\subsection{Clustering}
When only records satisfying certain criteria are need from a table, it is often unnecessary to scan the whole table. For example, if only records that occurred
in January 2010 are needed, then querying all the table would be an extremely unnecessary overhead. This is where clustering comes in. Clustering reflects how
the records are physically located together on disk. Records that are likely to be queried together will be stored physically together to avoid querying the
whole table. For the previously mentioned example, the data can be clustered by month, such that all the records of January 2010 are physically in the same
page, for example. In this case, only this page will need to be read. As
people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time.

@@

Given the different physical design methods explained in the previous section, we now look at how these methods apply to different types of information systems.
Mainly, how they apply to Disk-based relational database systems, main memory relational database systems, data warehouses, and XML databases. We mainly focus
on the first four techniques discussed in Section~\ref{SEC-ELEMENTS} as these are the main design decisions involved in physical design.

\subsection{Disk Based Relational Database System (DRDB)}
\label{SEC-DDRDB}

Relational Databases were first introduced by Codd~\cite{codd1970relational}. In relational databases, users specify the actions they want to execute without
having to worry about how these operations will actually be performed. This is where the role of physical design comes in. Physical design specifies the access
paths used to get data from the tables.

\sussubsection{Paritioning}

MySQL only supports horizontal partitioning.

@@

\section{Real Databases (incorporate in previous section)}

@@

@@

\bibliographystyle{plain}
\bibliography{references}

>>>>

\date{}
% make the title area
\maketitle

@@

Relational Database Systems have been the most popular type of Information Systems for decades. When we think of designing a database systems, we usually think
of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in database design.
All of our designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database performance. Physical
database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed. Data is stored in the
form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and access the data such that
queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing techniques,
partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques~\cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

This survey presents the different physical design patterns that have been used in different types of information systems. The systems discussed in this survey
are disk based relational databases (traditional database systems), main memory relational databases, data warehouses, and XML databases. The aim of this survey
is not to explain the details of every data structure used in physical design, but rather to compare how the different data structures have been used in these
systems. This includes explaining how alterations or additions are made to data structures to fit the specific needs of every system. For a complete reference
on the details of the data structures used in physical database design, please see (CITE HERE). Using the different physical design methods used in disk based
relational database systems, we explore how these methods apply to the other systems and where the differences and similarities lie.

The rest of the survey is organized as follows. Section~\ref{SEC-ELEMENTS} first explains different elements of physical database design based on traditional
disk-based databases. Section~\ref{SEC-DIFFSYS} then discusses how each of the three other systems uses these elements, as applicable, to improve its
efficiency.Sections~\ref{SEC-MMDB},~\ref{SEC-XML},~\ref{SEC-WAREHOUSES}, discuss Main Memory Databases, XML Databases, and Data Warehouses
respectively. Throughout these sections, comparisons are made as to how the different elements have been modified to fit the needs of each system.
Section~\ref{SEC-AUTO} then presents how the choice of the different physical design techniques can be automated. Section~\ref{SEC-OPEN} mentions some of the
open problems of physical database design, and suggests possible solutions. Finally Section~\ref{SEC-CONCL} summarizes and concludes this survey.

@@

Relational Databases were first introduced by Codd~\cite{codd1970relational}. Disk based Relational Database Systems (RDBMS) are the most common and popular
type of Information Systems. Since they have been extensively researched and have been around for many decades, we use them as the basis of our comparison. In
this section, we describe the different elements of physical design in RDBMS. These elements will also be examined in the other systems to see how they apply
to them.

@@

B-trees (short for Balanced Trees) are the most used database index structure.  The B-tree is essentially similar to the traditional Binary Search Tree, but
instead of havinsummarizesg one value per node, a B-tee can have many values per node~\cite{comer1979ubiquitous} as shown in Figure (PUT A FIGURE FOR A B TREE
HERE).
B-trees are suitable for relational databases since the cost of retrieval in a B-tree is at most proportional to: $log_{d}\frac{n+1}{2}$. The cost of insertion
and deletion is at most proportional to $log_{d}n$ due to the possibility of progressing back up the tree to balance it after an insertion or a deletion.
Although B-trees do well in retrieval, deletion and insertion, they do not perform well in sequential search.

@@

mention clustered vs. unclustered ?

@@

Despite the fact that materialized views provide a significant improvement in performance, we cannot simply create a materialized view for each common query. To
begin with, materialized views consume disk storage which might be limited. The second, and main, problem with materialized views is their maintenance. Ensuring
that the views have the most up to date data is tricky, and may outset the benefit of using materialized views if it is not properly designed. Finally, having
several materialized views may increase the cost of searching for the appropriate view to use during the query optimization stage. Data warehouses heavily rely
on materialized views since they have many queries with several joins. This will be discussed in details in Section~\ref{SEC-WAREHOUSES}.
There are many variations of B-trees. In relational databases, the most popular index is the $B^{+}-tree$. The $B^{+}-tree$ is a variant of the B  tree which is
becoming the main indexing method supported by many databases such as DB2, Oracle, and SQL Server~cite{lightstone2007physical}. The main difference between the
$B^{+}-trees$ and the B-tree is that only leaf nodes contain data pointers in a $B^{+}-tree$. (COMPARE COST HERE). In Prefix $B^{+}-trees$, instead of using the
actual key value of a record, a prefix from the key value is used to decrease the storage size needed which will in turn allow more records to be stored per
node thus decreasing the height of the tree.

@@

Put table from database about the diff horizontal partitioning 
MySQL only supports horizontal partitioning.

\subsection{Clustering}
When only records satisfying certain criteria are need from a table, it is often unnecessary to scan the whole table. For example, if only records that occurred
in January 2010 are needed, then querying all the table would be an extremely unnecessary overhead. This is where clustering comes in. Clustering reflects how
the records are physically located together on disk. Records that are likely to be queried together will be stored physically together to avoid querying the
whole table. For the previously mentioned example, the data can be clustered by month, such that all the records of January 2010 are physically in the same
page, for example. In this case, only this page will need to be read. As
people realized the potential of clustering, multidimensional clustering was introduced where data is clustered based on different criteria at the same time
(TALK MORE).

@@

Given the different physical design methods explained in the previous section, we now look at how these methods apply to different types of information systems.
Mainly, how they apply to Main Memory Relational Database Systems (MMDB), Data Warehouses, and XML databases. We mainly focus
on the first four techniques discussed in Section~\ref{SEC-ELEMENTS} as these are the main design decisions involved in physical design.

@@

@@

Choosing which indexes to create, which views to materialize, and how to partition data are hard choices. They are highly dependent on the schema used as well
as the query workload. The number of combinations of choices that can be made is very large making this task quite tedious and difficult for database
administrators. Accordingly, researchers have been inclined to find ways to automate these decisions. There were many research efforts done in this
area(E.g.~\cite{finkelstein1988physical, hammer76selec, frank1992selec}), but most of these efforts suffered from either examining too many configurations, not
accounting for the query workload, or not considering if the query optimizer will benefit from the proposed configurations or not. Chaudhuri and
Narasayya~\cite{chadhuri1997selec} were the first to propose an efficient automatic index selection tool that is dependent on the query workload and uses the
query optimizer to evaluate the proposed configuration. They find the best configuration for the workload by examining each query individually to produce
single-column indexes then combining these configurations. They then run a greedy algorithm to select the best configuration with the least cost (as determined
by the query optimizer). They then try to further optimize the configuration by looking for possible multicolumn indexes as a combination of the single-column
indexes already found. One of the main strengths of this technique was that all these indexes were virtually created such that the query optimizer could use
them, but without having to incur the cost of physically creating them. This solution was part of the research at Microsoft and was tested in Microsoft SQL
Server 7.0.

Similarly, IBM~\cite{schiefer1999db2, valentin2000select} suggested a similar technique for their database with some changes such as selecting the order of
columns within an index based on the order in which the query optimizer evaluates the predicates.

@@

\bibliographystyle{IEEEtran}
\bibliography{references}
